A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-05 tex/codatadp.tex
more

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment. In
the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
by using subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types $\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$, etc.

@@

  * Constructed examples
    * proving
    * model finding?

  * Isabelle benchmarks

@@

  * presented a decision procedure for datatypes and codatatypes

  * part of our program to reduce the gap between automatic and interactive
    theorem proving
  * and to enrich provers with more high-level features that can be implemented
    efficiently

Future work:

  * more evaluation
    * use it in a new Nitpick-like tool
  * recursive and corecursive functions
    (and rewriting)
  * extensions, perhaps straightforward, to polymorphism and nested
    (co)recursion

In principle, rank-1 (top-level)
polymorphism would be easy to support; nesting datatypes inside datatypes, and
likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion (as in the \ty{tree} example above) as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).

  * quantifiers?

\def\ackname{Acknowledgment}
\paragraph{\ackname.}

Clark Barrett and Cesare Tinelli.

Bosses: Viktor Kuncak, Stephan Merz, Tobias Nipkow, and Christoph Weidenbach.

The research was partially supported by the Deutsche
Forschungs\-gemein\-schaft (DFG) project
\relax{Hardening the Hammer} (grant Ni\,491\slash 14-1).

>>>>

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment.%
\footnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
would be easy to support; nesting datatypes inside datatypes, and
likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion (as in the \ty{tree} example above) as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
In the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
by using subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types $\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$, etc.

@@

@@

We presented a decision procedure for datatypes and codatatypes and implemented
it in the SMT solver CVC4. Our main theoretical contribution has been the
extension of the procedure with codatatypes. Codatatypes are not very difficult
per se, but there are tricky corner cases to take care of if we care about
completeness (which we do). On the practical side, we obtained interesting
results on Isabelle benchmarks, including some rather
%mind-boggling
puzzling bisimulation proofs.

This work is part of a larger program that aims at enriching automatic
provers with more high-level features and reducing the gap between automatic and
interactive theorem proving. As future work, we want to try out CVC4 and its
(co)datatypes for higher-order model finding in Isabelle, as an
alternative to the counterexample generator Nitpick
\cite{blanchette-nipkow-2010}. We also see some opportunities to enrich CVC4
with recursive and corecursive functions. Finally, it might be possible to go
further in terms of supporting nested and mixed (co)recursion and quantified
formulas over (co)datatypes in SMT solvers.

\def\ackname{Acknowledgment}
\paragraph{\ackname.}
We owe a great debt to Clark Barrett and Cesare Tinelli for datatype case blah
blah blah.
%
Our present and former bosses, Viktor Kuncak, Stephan Merz, Tobias Nipkow,
Cesare Tinelli, and Christoph Weidenbach, have either encouraged the research on
codatatype or at least tolerated it, both of which we are thankful for.
%
Blanchette's research was partially supported by the Deutsche
Forschungs\-gemein\-schaft (DFG) project
\relax{Hardening the Hammer} (grant Ni\,491\slash 14-1).
